---
title: "TBD"
title-block-banner: true
abstract: "TBD"
thanks: "Code and data supporting this analysis is available at: https://github.com/christina-wei/INF3104-3-US-Survey.git"
author: 
  - Michaela Drouillard
  - Christina Wei
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: ref.bib
---

# Story

Introduction - news, inflation, millennials and younger not doing better than their parents

\% of financial compare & % financial satisfaction - in 2021

Let's dig deeper into 2021

-   gender

-   education

-   Social class

-   age

How financial compare / financial satisfaction changed over the years - follow cohort and age

How does social class impact these perceptions? - link to Michaela's literature

# To Do

Christina

-   financial satisfaction

-   Social status

-   Simulated data

-   Data validation tests

-   README

Michaela

-   Supplementary survey: The survey should have an introductory section and include the details of a contact person. The survey questions should be well constructed and appropriate to the task. The questions should have an appropriate ordering. A final section should thank the respondent

-   Introduction

-   Discussions

-   Limitations & Future Work

## Financial Compare (old)

```{r}
#| message: false
#| echo: false
#| warning: false

# Reference code from https://sebastiansauer.github.io/percentage_plot_ggplot2_V2/
survey_data_2021 |>
  filter(!is.na(social_class1), !is.na(financial_compare)) |>
  ggplot(aes(x = financial_compare, group = social_class1)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  scale_y_continuous(labels=scales::percent) +
  ylab("Percentage of response") +
  xlab("Financial Compare") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(legend.position="none") +
  facet_grid(~social_class1)



```

```{r}
#| message: false
#| echo: false
#| warning: false

# Reference code from https://sebastiansauer.github.io/percentage_plot_ggplot2_V2/
survey_data_2021 |>
  filter(!is.na(family_income_bands), !is.na(financial_compare)) |>
  ggplot(aes(x = financial_compare, group = family_income_bands)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  scale_y_continuous(labels=scales::percent) +
  ylab("Percentage of response") +
  xlab("Financial Compare") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(legend.position="none") +
  facet_grid(~family_income_bands)



```

```{r}
#| message: false
#| echo: false
#| warning: false

# Reference code from https://sebastiansauer.github.io/percentage_plot_ggplot2_V2/
survey_data_2021 |>
  filter(!is.na(social_class), !is.na(financial_change)) |>
  ggplot(aes(x = financial_change, group = social_class)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  scale_y_continuous(labels=scales::percent) +
  ylab("Percentage of response") +
  xlab("Financial Change") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(legend.position="none") +
  facet_grid(~social_class) 
```

```{r}
#| message: false
#| echo: false
#| warning: false

# Reference code from https://sebastiansauer.github.io/percentage_plot_ggplot2_V2/
# survey_data_2016 |>
#   filter(!is.na(social_class), !is.na(financial_change)) |>
#   ggplot(aes(x = financial_change, group = social_class)) +
#   geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
#   scale_y_continuous(labels=scales::percent) +
#   ylab("Percentage of response") +
#   xlab("Financial Change") +
#   theme(axis.text.x = element_text(angle = 45, hjust=1)) +
#   theme(legend.position="none") +
#   facet_grid(~social_class) 
```
